Big Ideas in Sports Analytics and Statistical Tools for their Investigation

### 2022-09-28 response_to_reviewers.Rmd

> While Sections 2 and 3 are well done and complete, Sections 4 and 5 are weak and seem incomplete and unpolished.

>Likewise the Opportunities section should be expanded to include the proliferation of Sports Analytics conferences.

> My largest concern is that a cursory discussion of player contribution or player ratings is given and this seems a major oversight of one of the “big ideas” in sports analytics given the focus on this topic in many sports.  See, for example, Mark Glickman’s recent presentation on this at the 2022 Joint Statistical Meetings.  I don’t think that Hvattum (2019) is sufficient to cover the area of player contribution.

## Minor comments from Reviewer 1

> In the second sentence, baseball is introduced as a “naturally discrete” sport, but there are certainly many things about baseball which are not discrete, as talked about later in the article. It would be helpful to clarify what is meant by discrete in this setting.

We have added a paragraph clarify the distinction between "discrete" and "continuous" sports. In order to maintain the flow of the narrative, the paragraph does not appear here, but rather at the beginning of Section 2.

> Also, later on in the first paragraph, it would be useful to have a citation or two about the changes being brought about in baseball due to analytics.

We believe that the last sentence of the first paragraph does exactly that.

> In Section 2.1, Figure 1 (George Lindsey’s calculations) is introduced before the reader even knows who George Lindsey is.

We think this is mainly due to LaTeX floating the figure suboptimally, and this can be corrected in the page proofs. We have moved the sentence introducing Figure 1 and Table 1 until after the mention of Lindsey's work in the text to mitigate this problem.

> Furthermore, the expected run matrix is displayed “in its more familiar form”, but I would venture for most readers of this article, no form would be familiar.

We have replaced "more familiar" with "most common".

> I think the caption does explain the matrix nicely, although it might be useful to point out from which column in Lindsey’s table the matrix is derived.

Indeed, we note which column corresponds to the expected run matrix in the caption.

> In Figure 2, I found it quite surprising that the expected point value was so negative     at the 95 yard line. Were safeties very common in this era of American football? This might be worth mentioning/discussing.

XX: Greg or Quang to address?

> The title of Section 2.3 seems to indicate that the aforementioned sports have only discrete states, but this is not always the case. Some clarification would be useful here. In general, the organization of Section 2 is confusing. It first seems to be going sport by sport, but then 2.3 is about several different sports and then 2.4 is on a general topic altogether. As is, it reads somewhat disjointed and is hard to follow. Section 3 flows much better in this sense.

We thank the reviewer for this observation. We have reorganized the subsections of Section 2 with more explicit titles to improve the flow and readability of this Section.

> Table 2 seems unnecessary, as the content is entirely summarized in the text.

We have removed Table 2.

> After the comprehensive views in Sections 2 and 3, Section 4 seems sparse on details. Much detail is given on the Bradley-Terry model, but only one example is given of a possible usage. Has this been used in other sports? Has it changed how we think about those sports or team strategies? An explanation of Elo rating should be given; I for one have no knowledge of how to compute it.

We thank the reviewer for this suggestion. We have added a subsection on Elo rating.

XX

> There is a large body of literature on Bayesian state-space models in sports outside of the Glickman and Stern paper, mainly in soccer/association football. See for example Koopman and Lit (2015) and the various references in its introduction.

Thank you for this suggestion. We have added a reference to Koopman's paper.

> As mentioned earlier, Section 5 seems sparse on details, and furthermore it seems odd to choose chess as the case study when it has barely been mentioned previously in the article. I think this section could use a table of some sort, perhaps something that summarizes the CRAN task view at a high level. Another potential would be a figure which shows a typical workflow of sports analytics as explained in the text.

XX

> Figure 6 doesn’t add much to the text; I don’t think the reader better understands sports analytics by visualizing many team colors. Instead, it would be better to show  how the theming can lead to a more impactful graphic for sports, i.e. show how that package might be used in a real-world setting.

XX

> For Section 6, there is also now a “Big Data Derby” in horse racing.

We thank the reviewer for this suggestion. We have added a reference to it.

> All in all, I think the article explains the ideas well and covers a lot of ground, but more care should be taken towards the organization and what to include as figures. The last half of the paper has a very different pace than the first half. I think a common theme in the paper is the rise in tracking data and how that has impacted sports–perhaps that idea could be better unified and stated in a specific section of the article, or at the very least noted in the conclusion.

XX

> Page 4, line 56: Needs to be some sort of separation between the “absorbing states” and “touchdown”

We have added "absorbing states that include touchdowns..."

> Page 5, line 56: Typo for “causal”

Fixed.

> Page 5, line 57: Should be “too conservative”

Fixed.

> Page 7, line 37: Should be “led” not “lead”

Fixed.

### 2022-09-28 sports_analytics.Rmd
address #28; #30

## In baseball, the expected run matrix

For clarity, we start with baseball due to the discrete nature of the sport. 
In baseball, $s$ is typically determined by two factors: the configuration of the runners on base (there are 8 possibilities) and the number of outs (3 possibilities). 
Thus, there are $|S| = 24 = 8 \cdot 3$ basic states of an inning in baseball[^state], and we are often interested in the number of runs that will be scored from some state until the end of the inning. 
In this example using baseball, $\mathbb{E}[X | s] \,$ is the expected number of runs scored between now and the end of the inning given that the inning is currently in state $s$. 
The collection of estimates $\mathbb{E}[X | s]$ for all 24 states is called the *expected run matrix* [^matrix], and it is foundational in baseball analytics. 
Figure \ref{fig:lindsey} shows a reproduction of George @lindsey1963investigation's original calculations, and Table \ref{tab:erm} shows the expected run matrix in its more familiar form.

@@

Early work on this topic can be found in @lindsey1963investigation, who used play-by-play data to compute an empirical estimate for the mean number of runs scored in the remainder of the inning for each of these 24 possible states of an inning. 
This line of work led to analysis of all types of common baseball strategies. 
For example, many baseball teams elect to attempt a sacrifice bunt with a runner on first and no one out in the inning, with the goal of moving the runner to second base, at the cost of the batter being out. 
@Tango2007book (and many subsequent analyses) conclude that the sacrifice bunt is rarely worth it, because most teams would be expected to score more runs with a runner on first and no outs than they would with a runner on second and one out.

@@

## In American football {#sec:football}

@@

Other attempts at modeling expected points in football are @goldner2012markov and @goldner2017situational, who propose a Markov framework.
In particular, the author considers a football drive as an *absorbing Markov chain*, consisting of distinct *absorbing states* touchdowns, field goals, and other play outcomes. 
For any given play, the expected points are calculated using the absorption probabilities for different scoring events.

@@

### 4th down strategy

@@

Numerous other papers (see @lopez2020bigger for details) use the analysis of the expected number of points to improve fourth down strategy. In addition to @Romer2006, later work by @Yam2019 uses win probability (see Section \ref{sec:wp}), rather than expected points, and a casual inference framework to reach similar conclusions that NFL teams are to conservative in going for it on 4th down.  In addition, they estimate that a better strategy would be worth about 0.4 wins per season on average, a substantial amount in a season with only 16 games.

@@

## In sports with continuous states

@@

@@

See also @bornn2017studying for more information on how tracking data have enabled advanced statistical analyses of basketball in recent years.
The strategy of maximizing expected points in basketball has lead directly to the proliferation of three-point shooting in the NBA.

@@

@@

@kumagai2021hockey presents an EPV metric for ice hockey.

@@

For example, in the bottom of the ninth inning of a tied baseball game, the optimal strategy for winning the game is maximizing the probability of scoring *at least one run*, which may differ from the strategy of maximizing expected runs. 
If we let $U$ be the set of all strategies, then we assert that it is not always the case that the strategy $u \in U$ that maximizes the expected number of points will maximize the probability of winning:
$$
  \underset{u \in U}{\arg \max \,}{\Pr[X > 0 | s, u]} \neq \underset{u \in U}{\arg \max \,}{\mathbb{E}[X | s, u]} \,.
$$
Consider the situation where runners are on first and third base, and the score is tied in the bottom of the ninth inning with no one out. 
Table \ref{tab:tied} (derived from Table \ref{tab:erm}) reveals that the expected number of runs scored in the remainder of the inning is 1.94 runs, while the probability of scoring zero runs is 0.13. 
The defense is in a tight spot, facing an 87% probability of losing the game. 
However, by walking the hitter to load the bases, they create the opportunity to force the lead runner at home and thus reduce the chances of scoring to 82%, even though they raise the expected number of runs scored to 2.22. 
In this case, the *defensive* team is wise to pursue the strategy of maximizing the expected number of runs scored, because it *minimizes* the probability of scoring at least one run.

```{r}
erm <- erm |>
  mutate(
    Pr_0 = c(
      .747, .855, .933,
      .604, .734, .886, 
      .381, .610, .788,
      .120, .307, .738,
      .395, .571, .791,
      .130, .367, .717,
      .180, .270, .668,
      .180, .303, .671
    )
  )

@@

A third crucial idea in sports analytics is the estimation of team strength. 
A popular method for estimating team strength in sports is through pairwise evaluations. 
Perhaps the most famous probability model for predicting the outcome of a paired comparison is the Bradley-Terry model (BTM) [@Bradley1952]. 
For a pair of players (or teams) $i$ and $j$, let $\Pi_{ij}$ denote the probability that $i$ is preferred to $j$. 
Then the BTM is a logistic regression model with parameters $\beta_i, \beta_j$ such that
$$
\log\left(\frac{\Pi_{ij}}{\Pi_{ji}}\right) = \beta_i - \beta_j \,.
$$
Here, $\exp{\beta_i}$ is often viewed as a representation of team $i$'s ability.

@@

Another widely known tool for measuring team strength is the Elo rating system [@Elo1978], which was originally developed for chess, but also has been used to estimate team strength in other sports. 
See @koning2017rating for more information on applications of the Elo rating in soccer. 
Elo ratings are used extensively for rankings of teams in numerous sports by the data journalists at [FiveThirtyEight.com](https://fivethirtyeight.com).

@Glickman1998 propose a Bayesian state-space model for paired comparisons for predicting NFL games, allowing team strength parameters to vary over time. 
In particular, they model point differential in the NFL by introducing week-to-week and season-to-season as the two primary sources of variation in team strengths. 
See also @glickman2017estimating for more discussion on estimating team strengths in American football. 
More recently, @Lopez2018 extend @Glickman1998's state-space model to understand randomness in the four major American sports leagues. 
Betting moneylines are used in place of point differentials in order to estimate team strengths, and this framework also accounts for home advantage.
Both papers motivate the usefulness of model-based measures of team strength by demonstrating their superiority to low-resolution win-loss records. 
Apart from sports gambling, having an accurate estimate of team strength is useful to team officials, who are constantly monitoring and forecasting their team's ability.

>>>>

The concept of a state is easier to grasp in a sport that can be modeled as *discrete* (in the sense of [discrete event simulation](https://en.wikipedia.org/wiki/Discrete-event_simulation)). 
By discrete, we mean a sport that can be easily broken into short, distinct segments of action which are typically summarized categorically. 
Each of these segments might represent a state $s$. 
For example, each pitch in baseball is either a ball or a strike. 
If the ball is put in play, then there may be a complex sequence of movements by the players, but ultimately (within a few seconds) that sequence will end and no more action will be permitted until the next pitch. 
At the beginning and end of each phase of action, we will know definitively which team is on offense and defense, which runners are on which bases, the score, how many outs there are, etc. 
Tennis could similarly be viewed as a series of discrete actions defined by each point. 
To say that a player is winning 6-2, 3-1, 40-15 and serving with one fault committed is to characterize the state of the match.
In American football, the game can be broken down into a discrete sequence based on each down. 
Contrast this to sports like lacrosse, soccer, or any variant of hockey, which feature largely running clocks and continuous player movement. 
In these sports, it is not obvious how to break up the action into discrete chunks.

In this Section, we illustrate how the fundamental concept of the expected value of a game state leads to compelling findings in a variety of sports.

## Discrete event analysis

First, we explore results derived from the expected value of a state in sports where discrete event analysis is common. 
We draw primarily on baseball and American football, but applications in other sports (e.g., tennis) are common (see for example, @kovalchik2019calibration).

### In baseball, the expected run matrix

In baseball, $s$ is typically determined by two factors: the configuration of the runners on base (there are 8 possibilities) and the number of outs (3 possibilities). 
Thus, there are $|S| = 24 = 8 \cdot 3$ basic states of an inning in baseball[^state], and we are often interested in the number of runs that will be scored from some state until the end of the inning. 
In this example using baseball, $\mathbb{E}[X | s] \,$ is the expected number of runs scored between now and the end of the inning given that the inning is currently in state $s$. 
The collection of estimates $\mathbb{E}[X | s]$ for all 24 states is called the *expected run matrix* [^matrix], and it is foundational in baseball analytics.

@@

Early work on this topic can be found in @lindsey1963investigation, who used play-by-play data to compute an empirical estimate for the mean number of runs scored in the remainder of the inning for each of these 24 possible states of an inning. 
This line of work led to analysis of all types of common baseball strategies. 
For example, many baseball teams elect to attempt a sacrifice bunt with a runner on first and no one out in the inning, with the goal of moving the runner to second base, at the cost of the batter being out. Figure \ref{fig:lindsey} shows a reproduction of @lindsey1963investigation's original calculations, and Table \ref{tab:erm} shows the expected run matrix in its most common form.

@Tango2007book (and many subsequent analyses) conclude that the sacrifice bunt is rarely worth it, because most teams would be expected to score more runs with a runner on first and no outs than they would with a runner on second and one out.

@@

### In American football, expected points {#sec:football}

@@

Other attempts at modeling expected points in football are @goldner2012markov and @goldner2017situational, who propose a Markov framework.
In particular, the author considers a football drive as an *absorbing Markov chain*, consisting of distinct *absorbing states* that include touchdowns, field goals, and other play outcomes. 
For any given play, the expected points are calculated using the absorption probabilities for different scoring events.

@@

### In American football, 4th down strategy

@@

Numerous other papers (see @lopez2020bigger for details) use the analysis of the expected number of points to improve fourth down strategy. In addition to @Romer2006, later work by @Yam2019 uses win probability (see Section \ref{sec:wp}), rather than expected points, and a causal inference framework to reach similar conclusions that NFL teams are too conservative in going for it on 4th down.  In addition, they estimate that a better strategy would be worth about 0.4 wins per season on average, a substantial amount in a season with only 16 games.

@@

## Continuous event analysis

@@

### In basketball, expected point value

@@

See also @bornn2017studying for more information on how tracking data have enabled advanced statistical analyses of basketball in recent years.
The strategy of maximizing expected points in basketball has led directly to the proliferation of three-point shooting in the NBA.

### In American football, yards gained

@@

### In other sports

@@

@kumagai2021hockey offer an EPV metric for ice hockey.

@@

For example, in the bottom of the ninth inning of a tied baseball game, the optimal strategy for winning the game is maximizing the probability of scoring *at least one run*, which may differ from the strategy of maximizing expected runs. 
If we let $U$ be the set of all strategies, then we assert that it is not always the case that the strategy $u \in U$ that maximizes the expected number of points will maximize the probability of winning:
$$
  \underset{u \in U}{\arg \max \,}{\Pr[X > 0 | s, u]} \neq \underset{u \in U}{\arg \max \,}{\mathbb{E}[X | s, u]} \,.
$$
Consider the situation where runners are on first and third base, and the score is tied in the bottom of the ninth inning with no one out. 
Information derived from Table \ref{tab:erm} reveals that the expected number of runs scored in the remainder of the inning is 1.94 runs, while the probability of scoring zero runs is 0.13. 
The defense is in a tight spot, facing an 87% probability of losing the game. 
However, by walking the hitter to load the bases, they create the opportunity to force the lead runner at home and thus reduce the chances of scoring to 82%, even though they raise the expected number of runs scored to 2.22. 
In this case, the *defensive* team is wise to pursue the strategy of maximizing the expected number of runs scored, because it *minimizes* the probability of scoring at least one run.

```{r, eval=FALSE}
erm <- erm |>
  mutate(
    Pr_0 = c(
      .747, .855, .933,
      .604, .734, .886, 
      .381, .610, .788,
      .120, .307, .738,
      .395, .571, .791,
      .130, .367, .717,
      .180, .270, .668,
      .180, .303, .671
    )
  )

@@

A third crucial idea in sports analytics is the estimation of team strength. 
A popular method for estimating team strength in sports is through pairwise evaluations.

## Bradley-Terry models

Perhaps the most famous probability model for predicting the outcome of a paired comparison is the Bradley-Terry model (BTM) [@Bradley1952]. 
For a pair of players (or teams) $i$ and $j$, let $\Pi_{ij}$ denote the probability that $i$ is preferred to $j$. 
Then the BTM is a logistic regression model with parameters $\beta_i, \beta_j$ such that
$$
\log\left(\frac{\Pi_{ij}}{\Pi_{ji}}\right) = \beta_i - \beta_j \,.
$$
Here, $\exp{\beta_i}$ is often viewed as a representation of team $i$'s ability.

@@

## Elo ratings

Another widely known tool for measuring team strength is the Elo rating system [@Elo1978], which was originally developed for chess, but also has been used to estimate team strength in other sports. 
See @koning2017rating for more information on applications of the Elo rating in soccer, and @kovalchik2019calibration and @kovalchik2016searching for Elo ratings in tennis. 
Elo ratings are used extensively for rankings of teams in numerous sports by the data journalists at [FiveThirtyEight.com](https://fivethirtyeight.com).

## Bayesian state-space models

@Glickman1998 propose a Bayesian state-space model for paired comparisons for predicting NFL games, allowing team strength parameters to vary over time. 
In particular, they model point differential in the NFL by introducing week-to-week and season-to-season as the two primary sources of variation in team strengths. 
See also @glickman2017estimating for more discussion on estimating team strengths in American football.

More recently, @Lopez2018 extend @Glickman1998's state-space model to understand randomness in the four major American sports leagues. 
Betting moneylines are used in place of point differentials in order to estimate team strengths, and this framework also accounts for home advantage.
Both papers motivate the usefulness of model-based measures of team strength by demonstrating their superiority to low-resolution win-loss records. 
Apart from sports gambling, having an accurate estimate of team strength is useful to team officials, who are constantly monitoring and forecasting their team's ability.

@koopman2015dynamic uses a Bayesian state-space model to develop team ratings for English Premier League soccer that improves on bookmaker's odds.
